---
title: "New York Climate Budget Optimization Proof of Concept"
author: "Max Shron"
date: "10/7/2022"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lpSolve)
```

### Setup

Let's take a step back from the problem of optimizing a real budget, and instead talk about trying to optimize a simplified one with four programs.

Making up some program names, let's say we have HeatMid and HeatLow, which subsidize middle-income and low-income heat pumps respectively, and SolarMid and SolarLow, which do similar for rooftop solar. Two products, four programs, and for each we need to choose a subsidy level subject to an overall budget constraint.

Our objective is to maximize the green-house gas (GHG) savings caused by the money spent on each program, by and large CO2. How many incremental heat pumps and solar installations can we incentivize?

### Formalization

We have two products, so let's index each product with an integer (e.g. Heat Pump = 1, Solar Arrays = 2). We do similarly for each market (middle income households vs low income households). Then the subsidy $s_{ij}$ is the subsidy offered by a given program to subsidize good $i$ in market $j$. For example, if HeatMid offers a \$1000 subsidy, we'd say that $s_{ij}=1000$.

Let $q_{ij}(p)$ be the quantity of product $i$ which is purchased in market $j$ at price $p$. $q_{ij}$ is a function, whose parameters we need to estimate from some combination of data and heuristics. Keeping with a standard log-log form of the demand curve with constant price elasticity of demand, we assume that $q_{ij}(p) = X_{ij} p^{-\eta_{ij}}$. $\eta_{ij}$ is the price elasticity of demand for product $i$ in market $j$, and $X_ij$ is a scaling factor. In practice we would estimate these in log form, since $\log(q_{ij}(p) = \log(X_{ij} p^{-\eta_{ij}})=\log(X_{ij})-\eta_{ij}\log(p)$ which is conveniently linear.

In practice, we ought to be able to estimate the price elasticity of demand and the shift parameter empirically, using the subsides themselves as instrumental variables. The subsidies have changed periodically and without much warning, affecting the sales of the products and forming a kind of natural experiment on effective prices. For other markets, there are industry sources which publish estimates for the price elasticity of demand which we could access ourselves for money. And if we needed to, we could estimate the demand curve from Monte Carlo simulations of household behavior, which we have already taken a crack at in the case of heat pumps. Either way, once we have a demand curve we can get rolling.

For example, if 10,000 middle-income households bought \$10,000 heat pumps in one year under a \$2000 HeatMid subsidy, and 15,000 middle-income households bought \$10,000 heat pumps under a \$3000 subsidy, then:

```{r}
heatmid <- tibble(year = c(1,2), 
                  sticker_price = 10000, 
                  subsidy = c(2000, 3000),
                  price = sticker_price - subsidy,
                  qty = c(10000, 15000))

m.1 <- lm(log(qty) ~ log(price), data = heatmid)
m.1
```
Pulling out the coefficients,

```{r}
X_11 = exp(m.1$coefficients[1])
eta_11 = -m.1$coefficients[2]
```

We can then plot this (note that we're putting price on the x axis and quantity on the y axis, which is not normal in economics but more sensible to me and probably to other outsiders):

```{r}
# Define our quantity as a function of price, X and eta
qty <- function(X, eta, p) {X * p^(-eta)}

ggplot() +
  geom_function(fun = ~qty(X_11, eta_11, .x), xlim = c(3000,20000)) +
  xlab("Price") +
  ylab("Qty")
```

That is, at a price of $6,000 each, we expect `r qty(X_11, eta_11, 6000)` households to buy heat pumps next year.

Let $g_i$ be the greenhouse gas reduction from each unit of product $i$ which is purchased. Then we can draw a similar graph for GHG emission reduction in terms of price:

```{r}
g_1 = 10000 # kg of CO2 reduced per year per unit
ggplot() +
  geom_function(fun = ~ g_1 * qty(X_11, eta_11, .x), xlim = c(3000,20000)) +
  xlab("Price") +
  ylab("CO2 reduction")
```

So at a price of $6,000 per unit, we should expect `r g_1 * qty(X_11, eta_11, 6000)` kg of CO2 to be saved.

However, to understand the impact of the subsidy $s_{ij}$, we need to compare the quantity which will be bought with the subsidy vs the quantity which will be bought without the subsidy. 

To do this, we pick a base price $p_i$ to work off of. To continue this example of heat pumps for middle income households, let's say that our base price will be \$10,000, and our possible subsidy levels will run from 0 up to \$8,000. How much CO2 reduction can we expect at each subsidy level?

```{r}
p_1 <- 10000
ggplot() +
  geom_function(fun = ~ g_1 * (qty(X_11, eta_11, p_1 - .x) - qty(X_11, eta_11, p_1)), xlim = c(0,p_1*.8)) +
  xlab("Subsidy") +
  ylab("CO2 reduction")
```

As we get to an 80% subsidy, we save a ton of CO2, but it's also going to be extremely expensive to subsidize everyone's heat pump. What's the best use of our budget?

### Optimization

Let's say our total budget $B$ is \$1,000,000,000. How should we select all of the incentive levels so that we get the biggest GHG reduction for our given budget?

We start by expanding out the possible program subsidy levels and their resulting cost and GHG savings. Note again that all of these parameters are made up, but we could estimate them from real data.

```{r}
overhead = 1.2 # assume 20% overhead

s11 <- tibble(subsidy = "HeatMid",
              g = g_1,
              X = X_11,
              eta = eta_11,
              p = p_1,
              s = seq(100, p_1*.8, 1000), #max 80% subsidy
              name = paste(subsidy, s, sep="-"),
              q_nosub = qty(X, eta, p),
              q_sub = qty(X, eta, p - s),
              g_saved = g*(q_sub - q_nosub),
              s_total = s * q_sub * overhead)

s12 <- tibble(subsidy = "HeatLow",
              g = g_1,
              X = X_11*1.5, # larger market
              eta = eta_11*1.03, # more price sensitive
              p = p_1,
              s = seq(100, p_1*.8, 1000), #max 80% subsidy
              name = paste(subsidy, s, sep="-"),
              q_nosub = qty(X, eta, p),
              q_sub = qty(X, eta, p - s),
              g_saved = g*(q_sub - q_nosub),
              s_total = s * q_sub * overhead)

s21 <- tibble(subsidy = "SolarMid",
              g = 20000,
              X = X_11*0.5, # much smaller market
              eta = eta_11*0.93, # less price sensitive
              p = 10000,
              s = seq(100, 10000*.8, 1000), #max 80% subsidy
              name = paste(subsidy, s, sep="-"),
              q_nosub = qty(X, eta, p),
              q_sub = qty(X, eta, p - s),
              g_saved = g*(q_sub - q_nosub),
              s_total = s * q_sub * overhead)

s22 <- tibble(subsidy = "SolarLow",
              g = 20000,
              X = X_11*0.7, # much smaller market
              eta = eta_11*0.97,
              p = 10000,
              s = seq(100, 10000*.8, 1000), #max 80% subsidy
              name = paste(subsidy, s, sep="-"),
              q_nosub = qty(X, eta, p),
              q_sub = qty(X, eta, p - s),
              g_saved = g*(q_sub - q_nosub),
              s_total = s * q_sub * overhead)

s <- rbind(s11, s12, s21, s22)
B <- 1e9
```

And taking a look,
```{r}
s
```

Now we can use an integer programming package (`lpSolve`) to select the right levels of each program for us, subject to the constraint of the budget.

In this case, we have 32 possible combinations of (8 subsidy levels x 4 programs), so our linear solver will be a 32 dimensional Boolean vector $x\in{[0,1]}^{32}$.

To ensure that we only have one choice per level, they will be subject to the constraint that all of the Booleans for a given program must sum to 1 (so that, for example, $x_1 + x_2 + x_3 + x_4 + x_5 + x_6 + x_7 + x_8 = 1$).

We also want our budget to be within $B$, so we'll check that the linear combination $x\cdot s_{total} \leq B$, where $s_{total}$ is the total amount spent by the program at that subsidy level.

```{r}
num_programs <- s %>% count(subsidy) %>% nrow()

# goal: maximize ghg saved
f.obj <- pull(s, g_saved)

# subject to the following constraints
f.con <- matrix(c(s$s_total, # budget for each level
          ifelse(pull(s, subsidy) == "HeatMid", 1, 0), # must pick one level per program
          ifelse(pull(s, subsidy) == "HeatLow", 1, 0), # tk make this more automated
          ifelse(pull(s, subsidy) == "SolarMid", 1, 0),
          ifelse(pull(s, subsidy) == "SolarLow", 1, 0)),
          nrow=5, byrow = TRUE)

f.dir <- c("<=", replicate(num_programs, "="))
f.rhs <- c(B, replicate(num_programs, 1))
```

We run the optimization, using integer programming magic.

```{r}
optim.1 <- lp(direction = "max",
              f.obj,
              f.con,
              f.dir,
              f.rhs,
              all.bin = TRUE)
```

How close did we get to the budget, given these program options?

```{r}
sum(s$s_total[optim.1$solution == 1])/B
```

Which programs to fund, and at what subsidy level?

```{r}
s$name[optim.1$solution == 1]
```

How much millions of kg of GHG will we save per year?

```{r}
sum(s$g_saved[optim.1$solution == 1]) / 1e6
```
